---
title: "Wine Dataset"
date: "5/6/2024"
author: "STA 4990 Project"
output:
    pdf_document:
      df_print: kable
---

```{r}
knitr::opts_chunk$set(out.width = "50%", out.height="50%", fig.align="center", warning=FALSE, message=FALSE)

#Load in all the necessary packages for this data
library(carData)
library(caret)
library(ggplot2)
library(yardstick)
library(tidyr)
library(dplyr)
library(tidyverse)
library(mice)
library(glmnet)
library(GGally)
library(dslabs)
library(pROC)
library(mclust)
library(lubridate)
library(broom)

# Load in the wine data
setwd("C:/Users/Sonia/Desktop/Pau's Homework")
wine_df <- read.csv("wine-clustering.csv")  # kaggle's training set
wine_scaled <- as.data.frame(scale(wine_df))
wine_analysis <- wine_df
```

#EDA of Data

```{r}
summary(wine_df)
summary(wine_scaled)
```

```{r}
cor(wine_scaled)
```

## Boxplots of all data, to be honest not really needed but nice to have. summary does this well enough

```{r}
boxplot.matrix(as.matrix(select(wine_df, 1,10)))
boxplot.matrix(as.matrix(select(wine_df, 2,3,6,7,9,12)))
boxplot.matrix(as.matrix(select(wine_df, 8,11)))
boxplot.matrix(as.matrix(select(wine_df, 4)))
boxplot.matrix(as.matrix(select(wine_df, 5)))
boxplot.matrix(as.matrix(select(wine_df, 13)))
```

## Histograms for all variables

-   Risk had mentioned that their distributions could play a factor in helping kmeans choose the clusters. Ex is bimodal, does it separate them that way.

```{r}
par(mfrow=c(3,3))
par(bg="white")
for (i in 1:ncol(wine_df)){
hist(wine_df[[i]], main=paste("Histogram:", colnames(wine_df)[i]), xlab = paste(colnames(wine_df)[i],"Values"))
}


#applying log transformation
logMalic_Acid <- log(wine_df$Malic_Acid)
logNonflavonoid_Phenols<- log(wine_df$Nonflavanoid_Phenols)
logColor_Intensity <- log(wine_df$Color_Intensity)
logProline<- log(wine_df$Proline)

logtrans <- data.frame(logMalic_Acid, logNonflavonoid_Phenols, logColor_Intensity, logProline)
logtrans_scaled <- as.data.frame(scale(logtrans))

par(mfrow=c(3,3))
par(bg="white")
for (i in 1:ncol(logtrans_scaled)){
hist(logtrans_scaled[[i]], main=paste("Histogram:", colnames(logtrans_scaled)[i]), xlab = paste(colnames(logtrans_scaled)[i],"Values"))
}

wine_df_log <- data.frame(wine_df, logtrans)
wine_df_log <- select(wine_df_log, -Malic_Acid, -Nonflavanoid_Phenols, -Color_Intensity, -Proline)

wine_scaled_new <- data.frame(wine_scaled, logtrans_scaled)

wine_scaled_new <- select(wine_scaled_new, -Malic_Acid, -Nonflavanoid_Phenols, -Color_Intensity, -Proline)

#Proceeding with the new data wine_scaled_new with log transformations
cor(wine_scaled_new)
```

## PCA Scaled

```{r}
# finding number of significant PCs needed
prc_wine <- wine_scaled_new %>% prcomp(scale=TRUE)
pc_sd <- prc_wine$sdev
var_explained <- (pc_sd)^2 / sum(pc_sd^2)
plot(var_explained, xlab = "Principal Component", ylab = "Proportion of Variance Explained",
type = 'b', pch = 16, main = "Scree Plot")

PVE_table <- data.frame(var_explained,
cumsum(var_explained)
)
PVE_table

(abs(prc_wine$rotation[,1:7]) > 0.3)*prc_wine$rotation[,1:7] %>% round(4)
wine_augmented <- augment(prc_wine, wine_scaled_new)

p <- ggplot(wine_augmented, aes(col=Alcohol))
p1 <- p +geom_point(aes(x=.fittedPC1, y=.fittedPC2), size=2.5)
p1



p <- ggplot(wine_augmented, aes(col=Hue))
p2 <- p +geom_point(aes(x=.fittedPC1, y=.fittedPC2), size=2.5)
p2


p <- ggplot(wine_augmented, aes(col=Flavanoids))
p3 <- p +geom_point(aes(x=.fittedPC1, y=.fittedPC2), size=2.5)
p3

p <- ggplot(wine_augmented, aes(col=Total_Phenols))
p4 <- p +geom_point(aes(x=.fittedPC1, y=.fittedPC2), size=2.5)
p4

p <- ggplot(wine_augmented, aes(col=Hue))
p5 <- p +geom_point(aes(x=.fittedPC3, y=.fittedPC4), size=2.5)
p5

p <- ggplot(wine_augmented, aes(col=Ash))
p6 <- p +geom_point(aes(x=.fittedPC3, y=.fittedPC4), size=2.5)
p6
```

## The nonflavanoids enhance the colors in red wines, main component founded in white wines #this is all data from red wines \#

-   Need to name PCs!!!

```{r}
View(wine_augmented)


```

-   PC would be 2,3,4,6,8,12... most likely choose 2, 3, or 4

### Scree plot for clusters

```{r}

within_ss <- rep(NA, 13)
set.seed(4990) # it uses random initialization
for(i in 1:13){
k_mean <- kmeans(x=wine_scaled_new,
                    i,
                    nstart = 50)
within_ss[i] <- k_mean$tot.withinss
}
plot(1:13, within_ss, type = "b", ylab = "Objective Function", xlab = "# Clusters",
main = "Scree Plot")

```

-   number of clusters should be 2, maybe 3, possibly 11

### Creating clusters and assigning them to scaled set with PCs.

```{r}
set.seed(4990)
wine_cluster_3 <- kmeans(x=wine_scaled_new,
centers = 3,
nstart = 10)

cluster_3 <- factor
# assign clusters to augmented df
wine_augmented$clust_groups_3 <- as.factor(wine_cluster_3$cluster)
wine_augmented <- wine_augmented %>% select(-19:-27)

# unscaled clustering for k = 3 and assigning it
set.seed(4990)
wine_cluster_3_unscaled <- kmeans(x=wine_df_log,
centers = 3,
nstart = 10)
wine_augmented$clust_groups_3_unscaled <- as.factor(wine_cluster_3_unscaled$cluster)

View(wine_augmented)
```

## Now trying for cluster 2

```{r}
set.seed(4990)
wine_cluster_2 <- kmeans(x=wine_scaled_new,
centers = 2,
nstart = 10)
wine_augmented$clust_groups_2 <- as.factor(wine_cluster_2$cluster)

# unscaled clustering for k = 2 and assigning
set.seed(4990)
wine_cluster_2_unscaled <- kmeans(x=wine_df_log,
centers = 2,
nstart = 10)
wine_augmented$clust_groups_2_unscaled <- as.factor(wine_cluster_2_unscaled$cluster)

View(wine_augmented)

```

### Clustering for 11 groups

```{r}
#DO NOT DO THIS ONE, IT DOESNT MAKE SENSE

set.seed(4990)
wine_cluster_11 <- kmeans(x=wine_scaled,
centers = 11,
nstart = 20)
wine_augmented$clust_groups_11 <- as.factor(wine_cluster_11$cluster)
View(wine_augmented)
```

## cluster assignments between 2 and 3, 2nd table is distribution of 11 clusters

```{r}
table(wine_augmented$clust_groups_2, wine_augmented$clust_groups_3)

```

-   It is very interesting how they split the points. There must be a reason for the separation.

## PCA Graphs for k = 3

```{r}
ggplot(wine_augmented, aes(y = .fittedPC2, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_3)))
```

-   PC1 and PC2 seperate the clusters really well into distinct groups. significant variables for these PCs should be evaluated further.

-   different PC graphs, show PC1 and PC2 are the only important ones

```{r}
ggplot(wine_augmented, aes(y = .fittedPC3, x = .fittedPC2)) +
  geom_jitter(aes(colour = factor(clust_groups_3)))
ggplot(wine_augmented, aes(y = .fittedPC3, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_3)))
ggplot(wine_augmented, aes(y = .fittedPC4, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_3)))
ggplot(wine_augmented, aes(y = .fittedPC3, x = .fittedPC4)) +
  geom_jitter(aes(colour = factor(clust_groups_3)))
ggplot(wine_augmented, aes(y = .fittedPC2, x = .fittedPC4)) +
  geom_jitter(aes(colour = factor(clust_groups_3)))
```

### PCA Graphs for k = 2

```{r}
ggplot(wine_augmented, aes(y = .fittedPC2, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_2)))
```

-   Splits data pretty good

-   PCs show they like the k = 2 clustering as all of them show a clear decision between the values. More than k = 3.

```{r}
ggplot(wine_augmented, aes(y = .fittedPC3, x = .fittedPC2)) +
  geom_jitter(aes(colour = factor(clust_groups_2)))
ggplot(wine_augmented, aes(y = .fittedPC3, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_2)))
ggplot(wine_augmented, aes(y = .fittedPC4, x = .fittedPC3)) +
  geom_jitter(aes(colour = factor(clust_groups_2)))
ggplot(wine_augmented, aes(y = .fittedPC4, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_2)))
```

## PCA graphs for k = 11

```{r}
ggplot(wine_augmented, aes(y = .fittedPC2, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_11)))
ggplot(wine_augmented, aes(y = .fittedPC3, x = .fittedPC2)) +
  geom_jitter(aes(colour = factor(clust_groups_11)))
```

-   these are bad, clustering does not give a clear decision of the splitting.

# Unscaled PCA

### unscaled for k = 3 and k = 2

```{r}
ggplot(wine_augmented, aes(y = .fittedPC2, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_3_unscaled)))
ggplot(wine_augmented, aes(y = .fittedPC2, x = .fittedPC1)) +
  geom_jitter(aes(colour = factor(clust_groups_2_unscaled)))
```

-   Shows us that the variables need to be scaled as they are not on the same scales

# subset clusters

```{r}
cluster_k2_1 <- filter(wine_augmented, clust_groups_2 == "1")
cluster_k2_2 <- filter(wine_augmented, clust_groups_2 == "2")
cluster_k3_1 <- filter(wine_augmented, clust_groups_3 == "1")
cluster_k3_2 <- filter(wine_augmented, clust_groups_3 == "2")
cluster_k3_3 <- filter(wine_augmented, clust_groups_3 == "3")
```

# Perform PCA for each cluster

# Perform stat summary of each cluster

```{r}
cluster_k3 <- cbind(wine_analysis, wine_cluster_3$cluster)
cluster_k2 <- cbind(wine_analysis, wine_cluster_2$cluster)

k_3 <- factor(wine_cluster_3$cluster, order = TRUE,
levels = c(1:3))
k_2 <- factor(wine_cluster_2$cluster, order = TRUE,
levels = c(1:2))
cluster_k3 <- cbind(wine_analysis, k_3)
cluster_k2 <- cbind(wine_analysis, k_2)

# data for graphs
data_long <- cluster_k3 %>% 
  select(is.numeric, k_3)%>%
gather(var, value, -k_3)
ggplot(data_long, aes(group = k_3, y= value, fill = k_3), xlab="") +
geom_boxplot() +
labs(title="Numerical Boxplots by Cluster") +
facet_wrap(~var, scales="free_y", ncol=5) +
scale_y_continuous(
n.breaks = 10,
)


data_long <- cluster_k3 %>% 
  select(-1:-7, k_3)%>%
gather(var, value, -k_3)
ggplot(data_long, aes(group = k_3, y= value, fill = k_3), xlab="") +
geom_boxplot() +
labs(title="Numerical Boxplots by Cluster") +
facet_wrap(~var, scales="free_y", ncol=3) +
scale_y_continuous(
n.breaks = 10,
)

data_long <- cluster_k3 %>% 
  select(-7:-13, k_3)%>%
gather(var, value, -k_3)
ggplot(data_long, aes(group = k_3, y= value, fill = k_3), xlab="") +
geom_boxplot() +
labs(title="Numerical Boxplots by Cluster") +
facet_wrap(~var, scales="free_y", ncol=3) +
scale_y_continuous(
n.breaks = 10,
)

ggpairs_1 <- cluster_k2 %>% select(-7:-13)
names <- as.vector(colnames(ggpairs_1))
ggpairs(cluster_k2[,names]
        ,aes(col = as.factor(k_2)))
```
